---
title: "Chicago Budget"
author: "Michael Molloy"
date: "6/12/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r include=FALSE}
library(tidyverse)
```
#Chicago Budget Data 
The City of Chicago publishes dozens of datasets every year through their data portal (https://data.cityofchicago.org/).  Since 2011, they have made public the budgeted salaries/ hourly rates for all City of Chicago employees.

# Loading Source Data
Source data was exported from the Chicago data portal into CSV format. Each file was loaded into R as a dataframe. A "Year" column was added to keep track of which records come from which budget year. 

```{r}
# read data
budget_2011 <- read_csv("../input/source/chicago_budget/Budget_-_2011_Budget_Ordinance_-_Positions_and_Salaries.csv") %>% mutate(YEAR = 2011)
budget_2012 <- read_csv("../input/source/chicago_budget/Budget_-_2012_Budget_Ordinance_-_Positions_and_Salaries.csv") %>% mutate(YEAR = 2012)
budget_2013 <- read_csv("../input/source/chicago_budget/Budget_-_2013_Budget_Ordinance_-_Positions_and_Salaries.csv") %>% mutate(YEAR = 2013)
budget_2014 <- read_csv("../input/source/chicago_budget/Budget_-_2014_Budget_Ordinance_-_Positions_and_Salaries.csv") %>% mutate(YEAR = 2014)
budget_2015 <- read_csv("../input/source/chicago_budget/Budget_-_2015_Budget_Ordinance_-_Positions_and_Salaries.csv") %>% mutate(YEAR = 2015)
budget_2016 <- read_csv("../input/source/chicago_budget/Budget_-_2016_Budget_Ordinance_-_Positions_and_Salaries.csv") %>% mutate(YEAR = 2016)
budget_2017 <- read_csv("../input/source/chicago_budget/Budget_-_2017_Budget_Ordinance_-_Positions_and_Salaries.csv") %>% mutate(YEAR = 2017)
budget_2018 <- read_csv("../input/source/chicago_budget/Budget_-_2018_Budget_Ordinance_-_Positions_and_Salaries.csv") %>% mutate(YEAR = 2018)
budget_2019 <- read_csv("../input/source/chicago_budget/Budget_-_2019_Budget_Ordinance_-_Positions_and_Salaries.csv") %>% mutate(YEAR = 2019)
```
# Data Wrangling
Before any analysis can be performed on the dataset we need to tidy up the data. 

## Add Missing Columns
The first problem we encounter is mismatched and/or missing column names. Once we make it so all of the column names match, we can bind each year's dataset together into a single dataframe.
```{r}
# add columns to years where they're missing
  #2011
  budget_2011$`BARGAINING UNIT` <- NA
  names(budget_2011)[names(budget_2011)=="DEPARTMENT NAME"] <- "DEPARTMENT DESCRIPTION"
  names(budget_2011)[names(budget_2011)=="DIVISION NAME"] <- "DIVISION DESCRIPTION"
  names(budget_2011)[names(budget_2011)=="FUND NAME"] <- "FUND DESCRIPTION"
  names(budget_2011)[names(budget_2011)=="SECTION NAME"] <- "SECTION DESCRIPTION"
  names(budget_2011)[names(budget_2011)=="SCHEDULE"] <- "SCHEDULE / GRADE"
  names(budget_2011)[names(budget_2011)=="TOTAL BUDGETED UNITS"] <- "TOTAL BUDGETED UNIT"
  names(budget_2011)[names(budget_2011)=="SUBSECTION CODE"] <- "SUB-SECTION CODE"
  names(budget_2011)[names(budget_2011)=="SUBSECTION NAME"] <- "SUB-SECTION DESCRIPTION"
  budget_2011 <- within(budget_2011,rm("DEPARTMENT"))
  
  #2012
  names(budget_2012)[names(budget_2012)=="DEPARTMENT NUMBER"] <- "DEPARTMENT CODE"
  names(budget_2012)[names(budget_2012)=="SCHEDULE/GRADE"] <- "SCHEDULE / GRADE"
  
  #2011-2013
  budget_2011$`ORGANIZATION DESCRIPTION` <- NA
  budget_2012$`ORGANIZATION DESCRIPTION` <- NA
  budget_2013$`ORGANIZATION DESCRIPTION` <- NA
  
  # bind together all years into a single dataframe
budget <- rbind(budget_2011,budget_2012,budget_2013,budget_2014,budget_2015,budget_2016,budget_2017,budget_2018,budget_2019)

```


```{r}
# are department names consistent?
budget %>% 
  distinct(`DEPARTMENT DESCRIPTION`,`DEPARTMENT CODE`) %>% 
  group_by(`DEPARTMENT CODE`) %>% 
  summarise(count = n()) %>% 
  filter(count >1) %>% 
  ggplot(aes(reorder(`DEPARTMENT CODE`,count),count, fill = as.factor(count))) +
  geom_col() +
  guides(fill=FALSE)+
  coord_flip() + 
  labs(title = "Mismatched Department Names", subtitle = "Chicago Budget Data 2011-2019", x= "Department Code",y="# of Department Names")
```

```{r}
# make department names consistent
  ## load the mapping table
  DepartmentNameMap <- read.csv("./input/source/chicago_budget/Department Mapping Table.csv",stringsAsFactors = F)

  ## use the mapping table to replace the production data
    for (row in 1:NROW(DepartmentNameMap)) {
      from <- DepartmentNameMap[row,1]
      to <- DepartmentNameMap[row,2]
      budget$`DEPARTMENT DESCRIPTION` <- str_replace(budget$`DEPARTMENT DESCRIPTION`,from,to)
    }

# write the output file
write_rds(budget,"./input/source/chicago_budget/Chicago Budget.Rds")
```



# Chicago Budget Analysis

```{r}

#read data
budget <- read_rds("../input/source/chicago_budget/Chicago Budget.Rds")
```

```{r}
# calculate the total budget by year
budget_summary <- budget %>%
  group_by(YEAR) %>% 
  summarize(total =sum(`TOTAL BUDGETED AMOUNT`)/1000000000) %>% #calculate total budget and simplify
  mutate(growth = 100*(total - lag(total))/lag(total)) # how much did the budget grow from previous year?

# chart of total budget over time
budget_summary %>% 
  ggplot(aes(YEAR,total)) +
  geom_line() +
  scale_x_continuous(breaks = c(2011:2019)) +
 labs(title = "City of Chicago Budget",subtitle = "Employee Salaries", y="Dollars (in Billions)",x="")

# chart of year over year budget growth rate
budget_summary %>%
  ggplot(aes(YEAR,growth)) +
  geom_line() +
  geom_smooth(method="lm") +
  scale_x_continuous(breaks = c(2011:2019)) +
  labs(title = "City of Chicago Budget",subtitle = "Employee Salary Growth", y="Growth Rate(%)",x="")
```

```{r}
# total budget by department
budget %>%
  group_by(Department = `DEPARTMENT DESCRIPTION`) %>% 
  summarize(total = sum(`TOTAL BUDGETED AMOUNT`)) %>%
  arrange(desc(total)) %>%
  head(10) %>% 
  mutate(total = total/1000000000) %>% 
  ggplot(aes(reorder(Department,total),total, fill = as.factor(total))) +
  geom_col() +
  guides(fill=FALSE)+
  coord_flip() + 
  labs(title = "Top Departments", subtitle = "Total Budget Amounts 2011-2019", x= "",y="$(in billions)")
```


