---
title: "Chicago Budget"
author: "Michael Molloy"
date: "6/12/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r include=FALSE}
library(tidyverse)
```
#Chicago Budget Data 
The City of Chicago publishes dozens of datasets every year through their data portal (https://data.cityofchicago.org/).  Since 2011, they have made public the budgeted salaries/ hourly rates for all City of Chicago employees.

# Loading Source Data
Source data was exported from the Chicago data portal into CSV format. Each file was loaded into R as a dataframe. A "Year" column was added to keep track of which records come from which budget year. 

```{r}
# read data
budget_2011 <- read_csv("../input/source/chicago_budget/Budget_-_2011_Budget_Ordinance_-_Positions_and_Salaries.csv") %>% mutate(YEAR = 2011)
budget_2012 <- read_csv("../input/source/chicago_budget/Budget_-_2012_Budget_Ordinance_-_Positions_and_Salaries.csv") %>% mutate(YEAR = 2012)
budget_2013 <- read_csv("../input/source/chicago_budget/Budget_-_2013_Budget_Ordinance_-_Positions_and_Salaries.csv") %>% mutate(YEAR = 2013)
budget_2014 <- read_csv("../input/source/chicago_budget/Budget_-_2014_Budget_Ordinance_-_Positions_and_Salaries.csv") %>% mutate(YEAR = 2014)
budget_2015 <- read_csv("../input/source/chicago_budget/Budget_-_2015_Budget_Ordinance_-_Positions_and_Salaries.csv") %>% mutate(YEAR = 2015)
budget_2016 <- read_csv("../input/source/chicago_budget/Budget_-_2016_Budget_Ordinance_-_Positions_and_Salaries.csv") %>% mutate(YEAR = 2016)
budget_2017 <- read_csv("../input/source/chicago_budget/Budget_-_2017_Budget_Ordinance_-_Positions_and_Salaries.csv") %>% mutate(YEAR = 2017)
budget_2018 <- read_csv("../input/source/chicago_budget/Budget_-_2018_Budget_Ordinance_-_Positions_and_Salaries.csv") %>% mutate(YEAR = 2018)
budget_2019 <- read_csv("../input/source/chicago_budget/Budget_-_2019_Budget_Ordinance_-_Positions_and_Salaries.csv") %>% mutate(YEAR = 2019)
```
# Data Wrangling
Before any analysis can be performed on the dataset we need to tidy up the data. 

## Add Missing Columns
The first problem we encounter is mismatched and/or missing column names. Once we make it so all of the column names match, we can bind each year's dataset together into a single dataframe.
```{r}
# add columns to years where they're missing
  #2011
  budget_2011$`BARGAINING UNIT` <- NA
  names(budget_2011)[names(budget_2011)=="DEPARTMENT NAME"] <- "DEPARTMENT DESCRIPTION"
  names(budget_2011)[names(budget_2011)=="DIVISION NAME"] <- "DIVISION DESCRIPTION"
  names(budget_2011)[names(budget_2011)=="FUND NAME"] <- "FUND DESCRIPTION"
  names(budget_2011)[names(budget_2011)=="SECTION NAME"] <- "SECTION DESCRIPTION"
  names(budget_2011)[names(budget_2011)=="SCHEDULE"] <- "SCHEDULE / GRADE"
  names(budget_2011)[names(budget_2011)=="TOTAL BUDGETED UNITS"] <- "TOTAL BUDGETED UNIT"
  names(budget_2011)[names(budget_2011)=="SUBSECTION CODE"] <- "SUB-SECTION CODE"
  names(budget_2011)[names(budget_2011)=="SUBSECTION NAME"] <- "SUB-SECTION DESCRIPTION"
  budget_2011 <- within(budget_2011,rm("DEPARTMENT"))
  
  #2012
  names(budget_2012)[names(budget_2012)=="DEPARTMENT NUMBER"] <- "DEPARTMENT CODE"
  names(budget_2012)[names(budget_2012)=="SCHEDULE/GRADE"] <- "SCHEDULE / GRADE"
  
  #2011-2013
  budget_2011$`ORGANIZATION DESCRIPTION` <- NA
  budget_2012$`ORGANIZATION DESCRIPTION` <- NA
  budget_2013$`ORGANIZATION DESCRIPTION` <- NA
  
  # bind together all years into a single dataframe
budget <- rbind(budget_2011,budget_2012,budget_2013,budget_2014,budget_2015,budget_2016,budget_2017,budget_2018,budget_2019)

```

## Mismatched Department Names
Over time, the city of Chicago has changed the exact text they use to describe departments.  While the numerical department code has stayed the same, it would be helpful if the department name text was consistent in our dataset. 

The graphic below shows the number of unique department names. 

```{r}
# are department names consistent?
budget %>% 
  distinct(`DEPARTMENT DESCRIPTION`,`DEPARTMENT CODE`) %>% 
  group_by(`DEPARTMENT CODE`) %>% 
  summarise(count = n()) %>% 
  filter(count >1) %>% 
  ggplot(aes(reorder(`DEPARTMENT CODE`,count),count, fill = as.factor(count))) +
  geom_col() +
  guides(fill=FALSE)+
  coord_flip() + 
  labs(title = "Mismatched Department Names", subtitle = "Chicago Budget Data 2011-2019", x= "Department Code",y="# of Department Names")
```

To make it so the department names are consistent across years, we create a CSV file to map the original text and the text we want to change it to. Then a for loop is used to go through every record in the dataset to find and replace based on our mapping CSV file.
```{r}
# make department names consistent
  ## load the mapping table
  DepartmentNameMap <- read.csv("../input/source/chicago_budget/Department Mapping Table.csv",stringsAsFactors = F)

  ## use the mapping table to replace the production data
    for (row in 1:NROW(DepartmentNameMap)) {
      from <- DepartmentNameMap[row,1]
      to <- DepartmentNameMap[row,2]
      budget$`DEPARTMENT DESCRIPTION` <- str_replace(budget$`DEPARTMENT DESCRIPTION`,from,to)
    }
```

# Saving the Data
Once we've finished with our data wrangling we can save our tidy data to a single RDS file and begin our analysis.
```{r}
# write the output file
write_rds(budget,"../input/source/chicago_budget/Chicago Budget.Rds")
```

# Chicago Budget Analysis

```{r}

#read data
budget <- read_rds("../input/source/chicago_budget/Chicago Budget.Rds")
```

# Total Budget
```{r}
# calculate the total budget by year
budget_summary <- budget %>%
  group_by(YEAR) %>% 
  summarize(total =sum(`TOTAL BUDGETED AMOUNT`)/1000000000) %>% #calculate total budget and simplify
  mutate(growth = 100*(total - lag(total))/lag(total)) # how much did the budget grow from previous year?

# chart of total budget over time
budget_summary %>% 
  ggplot(aes(YEAR,total)) +
  geom_line() +
  scale_x_continuous(breaks = c(2011:2019)) +
 labs(title = "City of Chicago Budget",subtitle = "Employee Salaries", y="Dollars (in Billions)",x="")
```

# Top Departments
```{r}
# total budget by department
budget %>%
  group_by(Department = `DEPARTMENT DESCRIPTION`) %>% 
  summarize(total = sum(`TOTAL BUDGETED AMOUNT`)) %>%
  arrange(desc(total)) %>%
  head(10) %>% 
  mutate(total = total/1000000000) %>% 
  ggplot(aes(reorder(Department,total),total, fill = as.factor(total))) +
  geom_col() +
  guides(fill=FALSE)+
  coord_flip() + 
  labs(title = "Top Departments", subtitle = "Total Budget Amounts 2011-2019", x= "",y="$(in billions)")
```

# Job Titles
For every position there is a human readable job title ("Title Description") and a unique code ("Title Code").  We can group the records by Title Code to see which codes have multiple human readable names.  The histogram below shows that while most title codes have only one human readable job title, some have two or three. These will need to be cleaned up.
```{r}
budget %>% 
  distinct(`TITLE DESCRIPTION`,`TITLE CODE`) %>% 
  group_by(`TITLE CODE`) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(count)) +
  geom_histogram(bins = 3) +
  labs(title = "Mismatched Title Names", subtitle = "Chicago Budget Data 2011-2019", x= "Same Title Code",y="# of Title Names")
```
```{r}
budget %>% 
  distinct(`TITLE DESCRIPTION`,`TITLE CODE`) %>% 
  group_by(`TITLE CODE`) %>% 
  summarise(count = n()) %>% 
  filter(count >1) %>% 
  

```

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Load packages
library(stringdist)
library(reshape2)
library(stringr)
library(tm)

# select the field in the dataset with which you want to perform clustering
Field <- budget$`TITLE DESCRIPTION`

# create the stringdistmatrix using the Jaro–Winkler distance algorithm 
UniqueField <- unique(as.character(Field))

DistanceField <- stringdistmatrix(UniqueField,UniqueField,method = "jw") # you can experiment with changing the method to get better results

rownames(DistanceField) <- UniqueField

# loop the clustering algorithm until it reaches the desired average number of records. the thought being that if we have a dataset with records we expect to be mostly unique than the average records in a cluster will be low.  This loop will determine the right number of clusters for your dataset.

  # set starting number of clusters
  i <- 1
  
  # set a starting avg_cluster (should be greater than x)
  avg_cluster <- 2
  
  # this one is important. how many records do you expect to be similar?
  x <- 2
  
  # define variable for use in clustering algorithm
  hc <- hclust(as.dist(DistanceField))

  # begin the loop
  while (avg_cluster > x) {
     dfClust <- data.frame(UniqueField, cutree(hc, k=i))
     names(dfClust) <- c('UniqueField','cluster')
     avg_cluster <- mean(table(dfClust$cluster))
     i = i+1
     }

# compile the cluster data into a data.frame
t <- table(dfClust$cluster)
t <- cbind(t,t / length(dfClust$cluster))
t <- t[order(t[,2], decreasing=TRUE),]
p <- data.frame(factorName=rownames(t), binCount=t[,1], percentFound=t[,2])


dfClust <- merge(x=dfClust, y=p, by.x = 'cluster', by.y='factorName', all.x=T)

dfClust <- dfClust[rev(order(dfClust$binCount,dfClust$cluster)),] # sort by the size of the cluster bin then the cluster ID

names(dfClust) <-  c('cluster','UniqueField','binCount')
```


The relationship between the number of clusters and the total number of records is interesting. In theory, if a dataset has no discernable patterns with which to cluster on, then the number of clusters would be equal to the number of records. In other words, each record would be its own "cluster".
```{r echo=TRUE, message=FALSE, warning=FALSE}
# How many clusters did we end up with?
print(paste('Number of clusters:', i))

# How many records in the dataset?
print(paste('Number of records:',NROW(data)))
```

Now we can take a look at the top results to start identifying the "real" patterns and further refining our data.
```{r echo=TRUE, message=FALSE, warning=FALSE}
# print cluster results
kable(dfClust[order(dfClust[1:200,3], dfClust[1:200,2], decreasing = T),1:2]) %>%
  kable_styling(bootstrap_options = c("striped","hover", "condensed","responsive"),fixed_thead = T) %>%
  column_spec(1, bold = T, border_right = T) %>%
  column_spec(2, width = "5m")
```